How to Avoid OOTA Without Really Trying

### 2024-09-10 paper/ootavacuous.tex
Shrink section 9 (Conclusions) and combine with section 8
Section 9 includes a lengthy restatement of the contents of the paper.
This is not necessary since the reader has just encountered all of the
material.  Replace this restatement with a two-sentence summary and a
one-sentence conclusion, merged into section 8.

A final concluding sentence may need to be added.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\section{Future Directions}
\label{sec:Future Directions}

This paper focused solely on compilers, primarily those that do only
local per-thread anaylsis and optimization.
It introduced the notion of quasi-volatile object accesses and gave
a definition for semantic dependency that is relative to a particular
C++ implementation.

Further explorations could:
\begin{itemize}
\item	Extend these results to some classes of interpreters and JITs
	on one hand, and to link-time optimizations (LTO) on the other.
\item	Consider various classes of global analysis, such as
	demonstrating that a given expression will always evaluate
	to a constant.
\item	Examine the possibility of ``flattening'' optimizations that
	combine multiple threads into one.
\item	Explore cases in which semantic dependencies must be absolute,
	not relative to any particular compiler.
\item	Include the effects of input and output or other operations
	which can vary from one execution to another.
\item	Delineate more precisely the limits of permissible behavior
	for quasi-volatile object accesses.
\end{itemize}

\section{Conclusion}
\label{sec:Conclusion}

This paper has presented samples of code containing non-trivial
semantic dependencies, uncovering some shortcomings in typical definitions of
``semantic dependency''.
It pointed out examples where semantic dependencies
are a function of an execution rather than of the source code, and examples
where a single semantic dependency extends from multiple loads
to a store but not from any one of those loads.
These complex dependencies motivated a generalization of the definition
of ``OOTA cycle'', which is shown in
Section~\ref{sec:Semantic Dependencies Can Be Many-To-One}.

This generalization, combined with a focus on compiler-based loose C++
implementations using single-thread analysis, and a consideration of
the relation between source code and machine code, led to the formulation
of the Fundamental Property of semantic dependency shown in
Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
This in turn led to a precise definition of ``semantic dependency'' given in
Section~\ref{sec:For Compilers Using Single-Thread Analysis}.
This definition was used in
Section~\ref{sec:Verifying the Fundamental Property}
to demonstrate that these implementations are
incapable of producing OOTA cycles in UB-free programs,
provided they treat all nonvolatile atomic objects as
quasi volatile, obeying the limitations outlined in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.

A variant definition of ``semantic dependency'' given in
Section~\ref{sec:For Compilers Using Global Analysis}, appropriate for
compiler-based loose C++ implementations that don't restrict
their code transformations to those based on single-thread analysis,
was used to show that such implementations
are incapable of producing OOTA cycles in UB-free programs
provided they treat all atomic objects as volatile.

This work shows that avoiding OOTA cycles requires no changes to the
standard, to current implementions, or to user practices for portable
code.
Future implementations must avoid optimizations that invent, duplicate,
or repurpose atomic loads.
More work is required to delineate any further restrictions on future
implementations that might carry out optimizations based on global
inter-thread analysis.

>>>>

\section{Conclusions and Future Directions}
\label{sec:Conclusions and Future Directions}

This paper focused solely on compilers, primarily those that do only
local per-thread anaylsis and optimization.
It introduced the notion of quasi-volatile object accesses and gave
a definition for semantic dependency that is relative to a particular
C++ implementation.
It formulated a Fundamental Property of semantic dependency and proved
that implementations subject to some very mild restrictions on how
they treat atomic object accesses do have this property.
Finally, it demonstrated that satisfying the Fundamental Property
guarantees an implementation cannot give rise to OOTA executions.

The conclusion to be drawn from this work is that in this context,
avoiding OOTA cycles requires no changes to the standard,
to current implementions, or to user practices for portable code.
Further explorations could:
\begin{itemize}
\item	Extend these results to some classes of interpreters and JITs
	on one hand, and to link-time optimizations (LTO) on the other.
\item	Consider various classes of global analysis, such as
	demonstrating that a given expression will always evaluate
	to a constant.
\item	Examine the possibility of ``flattening'' optimizations that
	combine multiple threads into one.
\item	Explore cases in which semantic dependencies must be absolute,
	not relative to any particular compiler.
\item	Include the effects of input and output or other operations
	which can vary from one execution to another.
\item	Delineate more precisely the limits of permissible behavior
	for quasi-volatile object accesses.
\end{itemize}
